The relative roles of genetic, ecological and environmental factors in species radiations

### 2017-11-22 paper/draft.tex
comment m&ms
further questions to intro
write some intro

\section{Introduction}
island biota have always fascinated and inspired researchers (Darwin, Wallace, ...)

This interest in islands and their recognition as highly suitable model systems resulted in the formulation of
important theoretical frameworks, foremost MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB).
Later ... GDM.

These theories have in common a focus on species numbers and respective rates of migration, extinction and speciation.
While they are successful in describing species numbers, they are neutral in design (in the UNTB sense) and thus were not meant to consider species traits and compositions.

Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).

Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}

These patterns have been explained theoretically [here, here and here], but usually one at a time.
In the following, we aim to produce and thus explain these forementioned patterns using a single framework, and thus following the same set of (rather simple) rules.

Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
and what trait combinations constitute successful colonizer and radiating species, respectively.

[NEW ANGLE:]
island biota have always fascinated and inspired researchers (Darwin, Wallace, ...).
While often featuring unique species communities and harboring countless endemics,
islands are especially threatened by climate change, human impact and invasive species.

This interest in islands and their recognition as highly suitable model systems resulted in the formulation of
important theoretical frameworks, foremost MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB).
Later ... GDM.

Because the study of islands biogeography takes into account evolutionary time frames (millions of years),
comprehensive and meaningful data is difficult to obtain.
In a recent review \cite{leidinger2017biodiversity}, we analyzed a powerful alternative to field experiments in island biogeography
--- mechanistic (/simulation) models.
Unfortunately, most island models failed to include more explicit genetic or micro evolution.
Instead some models featured a concept known as ``protracted speciaton'' \cite{rosindellXXX}, where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this and genetic mechanisms have important ramifications for speciation/radiation and species richness.
[EXAMPLES!]

@@

In the following, we present a model that remedies both of the adressed shortcomings by implementing genetic processes on a niche-explicit background
in an island biogeography framework.
We will adress the questions of what relative impact does (a) genetic linkage and (b) reproductive isolation have on emerging species numbers and diversity,
and how (c) landscape heterogeneity (habitat niche diverstiy) and (d) topographical complexity differently affect radiatons.

@@

>>>>

\section{Introduction}
%% island biota have always fascinated and inspired researchers (Darwin, Wallace, ...)

%% This interest in islands and their recognition as highly suitable model systems resulted in the formulation of
%% important theoretical frameworks, foremost MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB).
%% Later ... GDM.

%% These theories have in common a focus on species numbers and respective rates of migration, extinction and speciation.
%% While they are successful in describing species numbers, they are neutral in design (in the UNTB sense) and thus were not meant to consider species traits and compositions.

%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
%% island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).

%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}

%% These patterns have been explained theoretically [here, here and here], but usually one at a time.
%% In the following, we aim to produce and thus explain these forementioned patterns using a single framework, and thus following the same set of (rather simple) rules.

%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
%% and what trait combinations constitute successful colonizer and radiating species, respectively.

%% [NEW ANGLE:]
Island biota have always fascinated and inspired researchers (Darwin, Wallace, ...). %TODO: refs
Nowadays islands are understood as well suitable model systems for ecology and evolution due to their well definable characteristics like isolation. %TODO: ref
However island biota are also increasingly at risk.
While often featuring unique species communities and harboring countless endemics, %TODO: refs
islands are especially threatened by climate change, human impact and invasive species. %TODO: refs
This makes understanding the functioning and evolution of island communities a high priority objective.

One of the first and to this day haighly influential formal theories of island biodiversity is MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB). %TODO: ref
In it, species richness on islands is the result of opposing rates of immigration and extinction, both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
This relationship was since successfully applied to a multitude of islands %TODO: refs
, but also island like systems, such as coral reefs, lakes/ponds or forest fragments. %TODO: refs
This general applicability, however, can not hide the fact that the ETIB fails to deliver a mechanistic explanation of the processes at play.
It has particularly been criticized for its lack of considering speciation. %TODO: refs

About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM). %TODO: ref
The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
whith islands emerging from sea level, increasing over time in size and topographic complexity, before eroding and subsiding in the ocean again.
In doing so, the model incorporates concepts of niche theory, speciation and species succession.
Yet, there are still many processes and drivers deemed important, which are not considered by any formal theory.

%%TODO: here come a list of processes with examples

The problem of investigating these processes is the time frame needed for a proper evaluation.
Because islands biogeography takes into account evolutionary time frames (millions of years),
comprehensive and meaningful data is difficult to obtain.
In a recent review \cite{leidinger2017biodiversity}, we analyzed a powerful alternative to field experiments in island biogeography
--- mechanistic (/simulation) models.
Unfortunately, most island models failed to include more explicit genetic or micro evolution.
Instead, some models featured a concept known as ``protracted speciaton'' \cite{rosindellXXX}, where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this and genetic mechanisms have important ramifications for speciation/radiation and species richness.
[EXAMPLES!]

@@

In the following, we present a model that remedies both of the adressed shortcomings by implementing genetic processes on a niche-explicit background
in an island biogeography framework.
We will adress the questions of what relative impact does (a) genetic linkage and (b) reproductive isolation have on emerging species numbers and diversity,
and how (c) landscape heterogeneity (habitat niche diverstiy) and (d) topographical complexity differently affect radiatons.
[further questions thus tackled: traits of colonizer species, ...]

@@

[SCENARIOS, REPLICATES]
